---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# biooracler

<!-- badges: start -->
<!-- badges: end -->

`biooracler` provides access to the [Bio-Oracle](https://bio-oracle.org/) dataset via ERDDAP server. It will partly replace [sdmpredictors](https://github.com/lifewatch/sdmpredictors).

## Installation

You can install the development version of biooracler from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("bio-oracle/biooracler")
```

## List Layers

You can look-up available layers or search specific environmental variables.

### Search available layers
```{r}
library(biooracler)
list_layers("Ocean Temperature")
```

### See all available layers
```{r}
list_layers()
```

### See all information of all layers
```{r}
list_layers(simplify = FALSE)
```


## Layer information
```{r}
info_layer("tas_baseline_2000_2020_depthsurf")
```


## Download layers

First you need to define dataset to download together with the time, latitude and longitude constrains

```{r}
dataset_id <- "tas_baseline_2000_2020_depthsurf"

time = c('2001-01-01T00:00:00Z', '2010-01-01T00:00:00Z')
latitude = c(10, 20)
longitude = c(120, 130)

constraints = list(time, latitude, longitude)
names(constraints) = c("time", "latitude", "longitude")
```

Define also the variables to download:
```{r}
variables = c("tas_max", "tas_min")
```

Perform download
```{r}
layers <- download_layers(dataset_id, variables, constraints)
```

You can modify the download path and the output type with the arguments `directory` and `fmt` respectively.

```{r}
dir <- tempdir()
download_layers(dataset_id, variables, constraints, fmt = "csv", directory = dir)


download_layers(dataset_id, variables, constraints, fmt = "raster", directory = dir)
```
