---
title: "Getting started with biooracler"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with biooracler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # The code below contacts the Bio-Oracle ERDDAP server, so it is not run when
  # the vignette is built (e.g. on CRAN). Run it yourself in an interactive
  # session to reproduce the output.
  eval = FALSE
)
```

## What is Bio-Oracle?

[Bio-Oracle](https://bio-oracle.org/) provides marine environmental layers
(temperature, salinity, nutrients, ice cover, and more) for both present-day
conditions and future climate-change projections. The data are served through an
[ERDDAP](https://erddap.bio-oracle.org/erddap/) server hosted by the Flanders
Marine Institute (VLIZ).

`biooracler` is a thin wrapper around
[`rerddap`](https://docs.ropensci.org/rerddap/) that makes it convenient to
search, inspect, and download these layers from R.

```{r setup}
library(biooracler)
```

## Listing and searching layers

Every dataset on the server has a unique `dataset_id`. Use `list_layers()` to see
what is available, or pass a free-text query to narrow the search.

```{r list}
# All layers
list_layers()

# Free-text search
list_layers("Ocean Temperature")
```

The naming of a `dataset_id` encodes the variable, time period, and depth, for
example `thetao_ssp585_2020_2100_depthsurf` (ocean temperature, SSP5-8.5 scenario,
2020-2100, surface). See `?dataset_id` for the full convention.

## Inspecting a single layer

`info_layer()` returns the metadata for one dataset, including its variables and
the ranges of its dimensions (time, latitude, longitude). Use it to find valid
constraint values before downloading.

```{r info}
info_layer("thetao_baseline_2000_2019_depthsurf")
```

## Downloading data

`download_layers()` is the main function. You give it a `dataset_id`, the
`variables` you want, and a list of `constraints` on the `time`, `latitude`, and
`longitude` dimensions.

```{r download}
dataset_id <- "thetao_baseline_2000_2019_depthsurf"

constraints <- list(
  time      = c("2000-01-01T00:00:00Z", "2010-01-01T00:00:00Z"),
  latitude  = c(40, 50),
  longitude = c(-10, 0)
)

layers <- download_layers(
  dataset_id,
  variables   = c("thetao_mean"),
  constraints = constraints
)

layers
```

By default the result is a `terra::SpatRaster`, ready for mapping and analysis:

```{r plot}
terra::plot(layers)
```

### Output formats

The `fmt` argument controls the return type:

- `"raster"` (default): a `terra::SpatRaster`.
- `"nc"`: the parsed NetCDF object from `rerddap`.
- `"csv"`: a `data.frame` of the values.

```{r fmt}
df <- download_layers(dataset_id, "thetao_mean", constraints, fmt = "csv")
head(df)
```

### Choosing where files are saved

Downloads are cached on disk. Pass `directory` to choose where, and `filename`
to give the file a human-readable name instead of the default cache hash.

```{r filename}
download_layers(
  dataset_id, "thetao_mean", constraints,
  directory = tempdir(),
  filename  = "thetao_baseline_NE_atlantic"
)
```

## A note on the `time` constraint

The `time` constraint is a **range**: every time step that falls within it is
returned as a separate layer. Decadal baseline datasets store snapshots on decade
boundaries, so a range that spans two snapshots returns two layers. To get a
single time step, set the start and end of `time` to the same value.

## Learn more

- `?download_layers`, `?list_layers`, `?info_layer` for function reference.
- `?dataset_id` for the dataset naming convention.
- The [Bio-Oracle website](https://bio-oracle.org/) for the science behind the data.
